---
title: "Gov 50 Final Project"
author: "Tristan Angieri"
description: "Psychedelic Research Project"
output:
  distill::distill_article:
    self_contained: false
---   



## Introduction: Context and Relevance

Exploring the patterns of psychedelic use offers critical insights into the human experience with these substances. This study aims to compare the age of initiation and frequency of psychedelic use between urban and rural areas in the U.S. in 2021. Understanding these patterns is pivotal for grasping the broader societal and psychological impacts of psychedelics, especially given the differences in accessibility and cultural attitudes towards these substances in various settings.

## Proposal

### Research Question

How do the age of initiation and frequency of psychedelic use differ between urban and rural areas in the U.S. in 2021, and how do these factors relate to the recency of use?

### Hypothesis

It is hypothesized that individuals in urban areas may initiate psychedelic use at a younger age than those in rural areas due to factors such as increased exposure and accessibility. Additionally, the frequency and recency of psychedelic use might correlate with the age of initiation, with earlier initiation potentially leading to more frequent use.

### Explanatory Variables of Interest:

#### Geographical Location (Rural vs. Urban):

Variable: RURURB00
Description: "RURAL OR URBAN SEGMENT"

#### Age of Initiation of Hallucinogen Use:

Variable: HALLUCAGE
Description: "AGE WHEN FIRST USED HALLUCINOGENS"

#### Frequency and Recency of Hallucinogen Use:

Variables: HALLUCYR (Frequency), HALLUCREC (Recency)
Descriptions: "HALLUCINOGEN - PAST YEAR USE" and "TIME SINCE LAST USED HALLUCINOGEN"

#### Control Variables:

Ever Used Hallucinogens: HALLUCFLAG
Age Last Used Hallucinogens: HALLUCAGLST

### Expected Observations

If data shows that individuals from urban areas initiate hallucinogen use at a younger age with higher frequency and recency compared to those from rural areas, it would support the hypothesis.

### Statistical Analysis

A regression analysis will be conducted with frequency and recency of hallucinogen use as dependent variables, and geographical location and age of initiation as main independent variables. The regression will control for factors such as ever using hallucinogens and age of last use.

### Data Source

The data for this study is sourced from the ["National Survey on Drug Use and Health 2021 (NSDUH-2021-DS0001)"](https://www.datafiles.samhsa.gov/dataset/national-survey-drug-use-and-health-2021-nsduh-2021-ds0001).

The National Survey on Drug Use and Health (NSDUH) is an annual survey conducted in the United States. It is one of the most comprehensive sources of data on substance use, including hallucinogens, and covers a wide demographic. NSDUH's methodology includes interviews with individuals aged 12 and older and spans various topics from frequency of drug use to mental health. This dataset is particularly suitable for this study as it provides detailed information on hallucinogen use patterns across different geographical locations. The richness and granularity of the data make it ideal for examining the nuances of urban and rural differences in psychedelic experiences.

The [dataset's specific variables](https://www.datafiles.samhsa.gov/sites/default/files/NSDUH_files/RDAS/RDAS_Comparability.xlsx) will provide a comprehensive perspective on the research question.

## Analysis
```{r}
library(dplyr)
nsduh <- load("./NSDUH_2021.RData")

#https://www.r-bloggers.com/2012/11/analyze-the-national-survey-on-drug-use-and-health-nsduh-with-r/
#http://asdfree.com/national-study-on-drug-use-and-health-nsduh.html
#https://github.com/ajdamico/lodown
# nsduh_tib <- as_tibble(nsduh)
# 
nsduh
```

```{r}
# PUF2021_100622
halluc_use <- PUF2021_100622 |>
  mutate(
    metro_level = case_when(
      COUTYP4 == 1 ~ "Large Metro",
      COUTYP4 == 2 ~ "Small Metro",
      COUTYP4 == 3 ~ "Non Metro"
    )
  ) |>
  select(
    # metro level
    metro_level,
    # respondent age
    AGE3,
    # population density
    PDEN10,
    # metro/non-metro
    COUTYP4,
    # ever used hallucinogens
    hallucflag,
    # PAST YEAR USE (0 did not use, 1 did use)
    hallucyr,
    # ever used hallucinogens (DMTAMTFXY, ECSTMOLLY, HALLUCOTH, KETMINESK, LSD, MESC, PCP, PEYOTE, PSILCY, SALVIADIV)
    hallucevr,
    # age of first hallucage
    hallucage
  )
```

```{r}
library(tidyverse)
```
### Age of First Hallucinogen Use

```{r}
filtered_halluc_use <- halluc_use |>
  filter(hallucage < 69)

summary_stats <- filtered_halluc_use |>
  summarise(
    Median = median(hallucage),
    Q1 = quantile(hallucage, 0.25),
    Q3 = quantile(hallucage, 0.75),
    Mean = mean(hallucage),
    SD = sd(hallucage),
    Min = min(hallucage),
    Max = max(hallucage)
  )

summary_stats

halluc_use_plot <- filtered_halluc_use |>
  ggplot(mapping = aes(x = hallucage)) +
  geom_histogram(binwidth = 1) +
  labs(
    title = "Age of First Hallucinogen Use",
    x = "Age of respondent at first Hallucinogen use (years)",
    y = "Count"
  ) +
  geom_vline(aes(xintercept = mean(hallucage)), color = "indianred") +
  geom_vline(aes(xintercept = median(hallucage)), color = "dodgerblue")

halluc_use_plot
```

This analysis reveals that the median age of first hallucinogen use among respondents is approximately 19 years, representing the midpoint of the distribution. Approximately 25% of respondents start using hallucinogens before or at age 17 (Q1), while another quarter initiate use at or after age 21 (Q3), indicating quartile boundaries. The mean age of initiation, around 19.65 years, closely aligns with the median, implying a relatively symmetrical distribution. The histogram closely resembles a normal distribution with a slight right skew, indicating that most respondents commence using hallucinogens in their late teens or early twenties, with a smaller fraction initiating use at older ages.
### Age of First Hallucinogen Use by Metrolevel

```{r}
halluc_use_box_plot <- filtered_halluc_use |>
  filter(
    hallucage < 69
  ) |>
  ggplot(
    mapping = aes(
      x = hallucage,
      fill = metro_level
    )
  ) +
geom_boxplot() +
  labs(
    title = "Age of First Hallucinogen Use by Metro Level",
    x = "Metro Level",
    y = "Age at first Hallucinogen use (years)"
  )

halluc_use_box_plot

summary_stats <- halluc_use |>
  filter(hallucage < 69) |>
  group_by(metro_level) |>
  summarize(
    Median = median(hallucage),
    Q1 = quantile(hallucage, 0.25),
    Q3 = quantile(hallucage, 0.75)
  )

summary_stats
```
The boxplot analysis reveals variations in the age of first hallucinogen use across different metro level areas. In Large Metro areas, the median age of initiation is around 19 years, while in Non Metro and Small Metro areas, it's about 18 years. The first quartile age (25th percentile) indicates that a quarter of individuals in Non Metro areas start using hallucinogens at a younger age, at approximately 16 years, compared to the other groups. The third quartile age (75th percentile) shows that the majority in all three areas initiate use by their early twenties, with Large Metro areas slightly later at around 22 years. These findings highlight regional differences in the timing of first hallucinogen use, with individuals in Large Metro areas tending to initiate use slightly later on average.

### Hallucinogen lifetime use and metro level

```{r}
fit_2 <- lm(COUTYP4 ~ hallucevr, data = halluc_use)

fit_2 |>
  broom::tidy() |>
  select(term, estimate) |>  
  knitr::kable(digits = 2)
```
The regression analysis exploring the impact of lifetime hallucinogen use on the type of county in which respondents reside yielded an interesting result. The coefficient for hallucevr, representing lifetime hallucinogen use, is 0.00. This implies that there is no statistically significant association between whether a respondent has ever used hallucinogens and the metro/non-metro status of their county. In other words, lifetime hallucinogen use does not appear to be a determining factor in differentiating respondents living in urban versus rural areas. This finding suggests that the use of hallucinogens is similarly prevalent across different types of geographical locations, contrary to what might be expected considering potential differences in accessibility and exposure.

The intercept of the regression, at 1.73, indicates the baseline level for the dependent variable (COUTYP4), which in this context could suggest the average metro/non-metro status when hallucinogen use is not considered. However, without further context on the coding of the COUTYP4 variable, the interpretation of this intercept may vary.

This result challenges common perceptions about the urban-rural divide in drug use patterns and highlights the need for a more nuanced understanding of the factors influencing hallucinogen use across different communities.
### Age of First Hallucinogen Use and Recent Use Probability

```{r}
halluc_use_filtered <- halluc_use |>
  filter(hallucage < 69)

logit_model <- glm(hallucyr ~ hallucage, data = halluc_use_filtered, family = "binomial")

summary(logit_model)

halluc_use_filtered$predicted_prob <- predict(logit_model, type = "response")

ggplot(halluc_use_filtered, aes(x = hallucage, y = predicted_prob)) +
  geom_point() +
  geom_smooth(
    method = "glm",
    method.args = list(family = "binomial")
  ) +
  labs(title = "Probability of Hallucinogen Use in the Past Year by Age of First Use",
       x = "Age of First Use",
       y = "Predicted Probability of Use in Past Year")

```
# Analysis Overview

The logistic regression model was fitted to examine the relationship between the age at first use of hallucinogens (`hallucage`) and the probability of using hallucinogens in the past year (`hallucyr`). Logistic regression was chosen for this analysis because the dependent variable, `hallucyr`, is binary (0 for did not use, 1 for did use), which aligns well with the logistic model's ability to predict the probability of a binary outcome based on an independent variable.

## Regression Results

### Coefficients:

- **Intercept**: The intercept (`-1.386519`) represents the log odds of using hallucinogens in the past year when the age of first use (`hallucage`) is 0. This is a theoretical baseline as the age of first use cannot be zero.
- **Age of First Use (`hallucage`)**: The coefficient for the age of first use is `0.008313`. This value indicates the change in the log odds of using hallucinogens in the past year for each additional year delay in the age of first use.

### Statistical Significance:

- The p-value for the age of first use is `0.0884`, marginally above the conventional threshold of `0.05` for statistical significance. This suggests a trend towards significance but does not provide strong evidence to conclusively state that the age of first hallucinogen use is a predictor of use in the past year.

### Model Fit:

- **Null Deviance vs. Residual Deviance**: The null deviance is reduced from `9542.2` to `9539.3` in the residual deviance with the inclusion of the `hallucage` variable, indicating that the model with the predictor provides a slightly better fit than the null model (a model with no predictors).
- **AIC (Akaike Information Criterion)**: The AIC of the model is `9543.3`. Lower AIC values generally indicate a better model, but without a comparative model, this value serves as a baseline for model comparison.

## Interpretation

The positive coefficient for `hallucage` suggests that there is a slight increase in the probability of using hallucinogens in the past year with each additional year in the age of first use. However, since the result is not statistically significant at the `0.05` level, this finding should be interpreted with caution. The marginal significance indicates that while there may be a relationship between the age of first use and the likelihood of using hallucinogens in the past year, the evidence is not strong enough to make a definitive conclusion. It is possible that other factors not included in the model might influence this relationship. The relatively small change in deviance suggests that the age of first use, while potentially informative, does not strongly predict hallucinogen use in the past year by itself.

# Conclusion

In conclusion, this analysis provides some indication that the age at which individuals first use hallucinogens might have a marginal effect on their likelihood of using hallucinogens in the most recent year. However, the evidence is not strong enough to assert a definitive causal relationship. Further research, possibly incorporating additional variables and a larger sample size, could provide more insights into this relationship.